---
title: "cleaning and wrangling data"
author: "Lena R Capece"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Cleaning and wrangling data practice

## Load packages
```{r load, warning = FALSE, message = FALSE}

library(dplyr)
library(tidyr)
library(readr)

## Attaching package: ‘dplyr’

# The following objects are masked from ‘package:stats’:
# 
#     filter, lag
# 
# The following objects are masked from ‘package:base’:
# 
#     intersect, setdiff, setequal, union

# use '::' to resolve masked shown above packages by making sure R is using the function from the specific package that you want e.g., dply::filter()


```

## Introduction
About the data:
This data was downloaded from [here](https://knb.ecoinformatics.org/view/df35b.304.2) on April 4, 2023. 

## Read data

```{r read, warning = FALSE, message = FALSE}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore data

```{r explore, eval= FALSE, echo = FALSE}
colnames(catch_original)

head(catch_original)

summary(catch_original)

sort(unique(catch_original$Year))

# want to know what regions are the data
(unique(catch_original$Region))

```

## Clean and reshape data

### Selecting and removing columns with 'select()'

```{r}

# catch_data <- catch_original %>% 
#   select(Region, Year, Chinook, Sockeye, Pink, Chum)

catch_data <-catch_original %>% 
  select(-All, -notesRegCode)

head(catch_data)

```


### Check column types

```{r}

dplyr::glimpse(catch_data)

## see that Chinook is a character when it should be numeric 'double'

```


### Changing column content using 'mutate()'

```{r changing data type}

catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)

```

### Checking NA

```{r checking NA}

i <- which(is.na(catch_clean$Chinook))

catch_data[i, ]

```


### Fixing Chinook column

```{r fixing chinook}

catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I",
         "1", 
         Chinook), 
         Chinook = as.numeric(Chinook))

## Check 
catch_clean[i,]

```


### Reshape data

From wide to long use 'pivot_longer()'
```{r reshape long}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), 
               names_to = "species",
               values_to = "catch")

head(catch_long)



```

From long to wide use 'pivot_longer()'

```{r reshape wide}

catch_wide <- catch_long %>% 
  pivot_wider(names_from = species, 
              values_from = catch)

head(catch_wide)

```

### Renaming columns 

```{r rename columns}

catch_long <- catch_long %>% 
  rename(catch_thousands = catch)

```

### Adding a column using 'mutate'

```{r rename mutate}

catch_long <- catch_long %>% 
  mutate(catch=catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)

```

### Split, Apply, Combine workflow

Here we are going to use the `group_by()` and `summarize()`
```{r }
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)

mean_reg_yr <- catch_long %>% 
  group_by(Region, Year) %>% 
  summarize(catch_mean = mean(catch))

head(mean_reg_yr)

n_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(n= n()) #number of rows for a given group 

head(n_region)


```
```{r filtering}

sse_catch <- catch_long %>% 
  filter(Region == "SSE") #SSE is a character so need to use quotations

head(sse_catch)

```


```{r putting it all together!!!}
  
mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch))

head(mean_region)

```



