---
title: "Intro to Data Visualization"
author: "Lena R Capece"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load packages, message=FALSE, warning=FALSE}

library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)

```

## Load data

```{r load data, message=FALSE, warning=FALSE}

escapement <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")

```

## Explore data

```{r explore data, message=FALSE, warning=FALSE, echo=FALSE}

colnames(escapement)

## Peak at each column and class
glimpse(escapement)

## From when to when was data collected
range(escapement$sampleDate)

## How frequently was data collected
head(escapement$sampleDate) # begining of data
tail(escapement$sampleDate) # end of data

## Which species were collected?
unique(escapement$Species)

```
## Cleaning and wrangling data for data visualizing
### Calculate an annual escapement by SASAP.Region, Species and Species type

```{r cleaning, message=FALSE, warning=FALSE, echo=FALSE}

annual_esc <- escapement %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% #isolate temporal collection data by year, month, and day -- they are characters
  mutate(Year = as.numeric(Year)) %>% # make all "Year" a numeric because it is a character
  group_by(Species, SASAP.Region, Year) %>% # we want to group data by species SASAP. Region and Year
  summarize(escape = sum(DailyCount)) %>% # apply a single calculation to DailyCount to the entire dataset so we get escape counts for the entire year
  filter(Species %in% c("Chinook", "Sockeye","Chum","Coho","Pink")) # only get results for the intended species which are as rows

unique(annual_esc$Species)
head(annual_esc)


```

## Plotting basics with `ggplot`

```{r plotting essentials, message=FALSE, warning=FALSE,}
## Option 1
ggplot(data = annual_esc,
       aes(x = Species,
           y = escape)) +
  geom_col()
  
## Option 2
ggplot(data = annual_esc) + 
  geom_col(aes(x=Species, y=escape))

## Option 3
ggplot() +
  geom_col(data= annual_esc,
           aes(x=Species, y=escape))

```

## Different geoms

*Boxplot*
```{r}

annual_esc %>% 
  filter(Year == 1974, Species %in% c("Chum", "Pink")) %>% 
  ggplot(aes(x=Species, y=escape)) +
  geom_boxplot()
  

```
*Line and point*

```{r}

annual_esc %>% 
  filter(Species == "Sockeye") %>% 
  filter(SASAP.Region == "Bristol Bay") %>% 
  ggplot(aes(x=Year, y=escape)) +
  geom_line() +
  geom_point()

```

## Customization of plots 

*Basic plot*
```{r}
# basic plot 
ggplot(data=annual_esc, aes(x=Species, y=escape)) +
  geom_col()
  

```

*Make a column blue*
```{r}
# now make a column blue 
ggplot(data=annual_esc, aes(x=Species, y=escape, fill="blue")) +
  geom_col()

```
This didnt work--why? ggplot thinks fill is a variable. Instead we should add the color to the geom_col(). 

*Correct version of making column blue*
```{r}

ggplot(data=annual_esc, aes(x=Species, y=escape)) +
  geom_col(fill="blue")

## Yay!

```

*Color the columns by a different variable that is provided in the dataset such as SASAP.Region*

```{r}
## Put the condition and variablee inside the aes() function 

ggplot(data=annual_esc, aes(x=Species, y=escape, fill=SASAP.Region)) +
  geom_col()

```


## Creating multiple plots 
*Subset data* 

```{r}

annual_esc_2000 <- annual_esc %>% 
  filter(Year %in% c(2000:2016))

#check data 
unique(annual_esc_2000$Year)


```

*Plot with facet wrap*

```{r}

ggplot(annual_esc_2000, 
       aes(x=Year, y= escape, color = Species)) +
      geom_line() +
      geom_point() +
      facet_wrap(~SASAP.Region,
                scales = "free_y") #~ create the faceting by that variable

```


*Setting ggplot themes*

```{r}

ggplot(annual_esc_2000, 
      aes(x=Year, y= escape, color = Species)) +
      geom_line() +
      geom_point() +
      facet_wrap(~SASAP.Region,
                scales = "free_y") +
      theme_light() + 
      ylab("Escapement") +
      ggtitle("Annual Salmon Escapemeny by Region") +
      theme(plot.title = element_text(hjust = 0.5))


```

*Removing legend title and moving to bottom of plot*

```{r}

ggplot(annual_esc_2000, 
      aes(x=Year, y= escape, color = Species)) +
      geom_line() +
      geom_point() +
      facet_wrap(~SASAP.Region,
                scales = "free_y") +
      theme(legend.position = "bottom", legend.title = element_blank())

```
*Saving theme to object*
```{r}

my_theme <- theme_bw() +
  theme(legend.position = "bottom", legend.title=element_blank())

ggplot(annual_esc_2000, 
      aes(x=Year, y= escape, color = Species)) +
      geom_line() +
      geom_point() +
      facet_wrap(~SASAP.Region,
                scales = "free_y") +
      ggtitle("Annual Salmon Escapemeny by Region") +
      my_theme


```

*Make x axis labels at 45 degree*

```{r}

my_theme <- theme_bw() +
  theme(legend.position = "bottom", legend.title=element_blank())

ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))
```
*Smarter ticks using `scales()`*

```{r}

ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    scale_y_continuous(labels = comma) +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

```

*Make your plot an object*

```{r}

annual_region_plot <- ggplot(annual_esc_2000,
      aes(x = Year,
           y = escape,
           color = Species)) +
      geom_line() +
      geom_point() +
      scale_x_continuous(breaks = seq(2000, 2016, 2)) +
      scale_y_continuous(labels = comma) +
      facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
      ylab("Escapement") +
      ggtitle("Annual Salmon Escapement by Region") +
      my_theme +
      theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

annual_region_plot
```
## Saving plot

```{r}

ggsave("figures/annual_region_plot.png", annual_region_plot, width = 12, height = 8, unit="in")
```



## Reordering things 
*Number Years of data for each salmon species at Kuskokwim*
```{r}
n_years <- annual_esc %>%
    group_by(SASAP.Region, Species) %>%
    summarize(n = n()) %>%
    mutate(n = as.numeric(n))
    filter(SASAP.Region == "Kuskokwim")

# ggplot(n_years, aes(x=Species, y=n)) +
#   geom_bar()


class(n_years$n)
```

## Maps with leaflet
*Subset location information*
```{r}

location <- escapement %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()

```

*Create a map with leaflet*
```{r}

leaflet(location) %>% 
  addTiles() %>% 
  addMarkers(lng=~Longitude,lat=~Latitude, popup = ~Location) 

```


```{r}

leaflet(location) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers(
        lng = ~ Longitude,
        lat = ~ Latitude,
        popup = ~ Location,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1
    )
```

