---
title: "Intro to Data Visualization"
author: "Lena R Capece"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load packages, message=FALSE, warning=FALSE}

library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)

```

## Load data

```{r load data, message=FALSE, warning=FALSE}

escapement <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")

```

## Explore data

```{r explore data, message=FALSE, warning=FALSE, echo=FALSE}

colnames(escapement)

## Peak at each column and class
glimpse(escapement)

## From when to when was data collected
range(escapement$sampleDate)

## How frequently was data collected
head(escapement$sampleDate) # begining of data
tail(escapement$sampleDate) # end of data

## Which species were collected?
unique(escapement$Species)

```
## Cleaning and wrangling data for data visualizing
### Calculate an annual escapement by SASAP.Region, Species and Species type

```{r cleaning, message=FALSE, warning=FALSE, echo=FALSE}

annual_esc <- escapement %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% #isolate temporal collection data by year, month, and day -- they are characters
  mutate(Year = as.numeric(Year)) %>% # make all "Year" a numeric because it is a character
  group_by(Species, SASAP.Region, Year) %>% # we want to group data by species SASAP. Region and Year
  summarize(escape = sum(DailyCount)) %>% # apply a single calculation to DailyCount to the entire dataset so we get escape counts for the entire year
  filter(Species %in% c("Chinook", "Sockeye","Chum","Coho","Pink")) # only get results for the intended species which are as rows

unique(annual_esc$Species)
head(annual_esc)


```

## Plotting basics with `ggplot`

```{r plotting essentials, message=FALSE, warning=FALSE,}
## Option 1
ggplot(data = annual_esc,
       aes(x = Species,
           y = escape)) +
  geom_col()
  
## Option 2
ggplot(data = annual_esc) + 
  geom_col(aes(x=Species, y=escape))

## Option 3
ggplot() +
  geom_col(data= annual_esc,
           aes(x=Species, y=escape))

```

## Different geoms

*Boxplot*
```{r}

annual_esc %>% 
  filter(Year == 1974, Species %in% c("Chum", "Pink")) %>% 
  ggplot(aes(x=Species, y=escape)) +
  geom_boxplot()
  

```
*Line and point*

```{r}

annual_esc %>% 
  filter(Species == "Sockeye") %>% 
  filter(SASAP.Region == "Bristol Bay") %>% 
  ggplot(aes(x=Year, y=escape)) +
  geom_line() +
  geom_point()

```

## Customization of plots 

*Basic plot*
```{r}
# basic plot 
ggplot(data=annual_esc, aes(x=Species, y=escape)) +
  geom_col()
  

```

*Make a column blue*
```{r}
# now make a column blue 
ggplot(data=annual_esc, aes(x=Species, y=escape, fill="blue")) +
  geom_col()

```
This didnt work--why? ggplot thinks fill is a variable. Instead we should add the color to the geom_col(). 

*Correct version of making column blue*
```{r}

ggplot(data=annual_esc, aes(x=Species, y=escape)) +
  geom_col(fill="blue")

## Yay!

```

*Color the columns by a different variable that is provided in the dataset such as SASAP.Region*

```{r}
## Put the condition and variablee inside the aes() function 

ggplot(data=annual_esc, aes(x=Species, y=escape, fill=SASAP.Region)) +
  geom_col()

```


