---
title: 'R Practice: Tidy Data and Joins'
author: "Lena R Capece"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages 

```{r load packages, warning=FALSE, message=FALSE}

library(readr)
library(dplyr)
library(lubridate)

```

## Download data

```{r download data, warning=FALSE, message=FALSE }

bird_observations <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=53edaa7a0e083013d9bf20322db1780e")
surveys  <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=b2466fa5cb5ed7ee1ea91398fc291c59")
sites <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=81bf72420e69077097fb0790dcdc63a6")
taxalist <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-cap.256.10&entityid=58f863b7e3066e68536a9cacdc7bd58e")

```

## 2a. Explore data 

```{r warning=FALSE, message=FALSE, eval=FALSE, echo=FALSE}

colnames(bird_observations)

head(bird_observations)

summary(bird_observations)

```
## 2b. What are the primary and foreign keys for the tables bird_observations and taxalist?
*bird_observations*
foreign keys: species ID
primary keys

*taxalist*
foreign keys: species ID
primary keys

## Subset birds

```{r warning=FALSE, message=FALSE}
birds_subset <- bird_observations %>% 
      filter(species_id %in% c("BHCO", "RWBL")) %>% 
      filter(site_id %in% c("LI-W", "NU-C"))

```

## Join tables using left_join()

Joining birds_subset with sites will create wide dataset which will add columns from sites to birds_subset that correspond to the appropriate site ID. 


```{r warning=FALSE, message=FALSE}

birds_left <- birds_subset %>% 
  left_join(y = sites, by = "site_id")

```

