---
title: "RMarkdown_Analysis"
author: "Lena R Capece"
date: "2023-04-03"
output: 
  html_document:
    toc: true
    toc_float: true

---

## Introduction
This data set contains North Pole Environmental Observatory Bottle Chemistry data. This data set includes diverse measurements of seawater chemistry along with supporting conductivity, temperature, and depth (CTD) data from the Arctic Ocean near the North Pole.

Datas was downloaded on April 3, 2023 from [here.](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X)


### Set up

```{r setup read data, warning=FALSE, message=FALSE, echo=FALSE}

library(readr)


```


```{r read data, echo=FALSE, warning=FALSE, message=FALSE}

bg_chem <- read_csv("data/BGchem2008data.csv") 

#readr::read_csv("data/BGchem2008data.csv") tells us to read the file using the package
```

#### Data exploration

```{r data exploration, eval=FALSE, echo=FALSE}
## to preview data

##column names
colnames(bg_chem)

## structure data
str(bg_chem)

## summary stats
summary(bg_chem)

### min, median, mean and max of columns of data

## open data in a tab
#View(bg_chem)

##unique values in a column
unique(bg_chem$Date)


```


## Analysis

### Calculate summary statistics

```{r mean calc}

nitrate <- mean(bg_chem$NO3)
nitrite <-mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)
```


### Calculate and mean Redfield ratio

Redfield ratio 16:1
```{r ratio calc}

ratio <- (amm+nitrate+nitrite)/phos

```


### Plot Redfield ratio

```{r}

plot(bg_chem$P, bg_chem$NO3+bg_chem$NO2+bg_chem$NH4)

```


## Conclusion

The Redfield ratio is `r round(ratio)`. 
